---
title: "ADA_FINAL"
author: "Agokey"
date: "2024-11-24"
output: html_document
---
```{r} 
#Install packages
library(tidyverse)
library(table1)
library(haven)
library(magrittr)
library(ggfortify)
library(dplyr)
library(ggplot2)
library(gmodels)
library(grid)
library(png)
library(MASS)
library(haven)
library(labelled)
library(foreign)
library(DiagrammeR)
```

```
```
```{r}
# Load necessary libraries
library(haven)  # For reading SAS XPT files
library(dplyr)  # For data manipulation
```


```{r}
# Read the datasets(Datasets based on demographic,24-hr dietary recall, sex hormones and BMI)
demo <- read_xpt("C:/Users/percy/Downloads/DEMO_I(1).XPT")
foodint <- read_xpt("C:/Users/percy/Downloads/DR1IFF_I.XPT")
hormone <- read_xpt("C:/Users/percy/Downloads/TST_I.XPT")
BMI <- read_xpt("C:/Users/percy/Downloads/BMX_I(1).XPT")
```

```{r}
# Display column names for each dataset
colnames(demo)
colnames(foodint)
colnames(hormone)
colnames(BMI)
```
```{r}
# Find shared columns across datasets
common_cols <- Reduce(intersect, list(colnames(demo), colnames(foodint), colnames(hormone), colnames(BMI)))
common_cols  # Print shared columns
```

```{r}
# Merge the datasets
merged_data <- demo %>%
  full_join(foodint, by = "SEQN") %>%
  full_join(hormone, by = "SEQN") %>%
  full_join(BMI, by = "SEQN")
```

```{r}
library(dplyr)

# Aggregate demographic, hormone, or dietary data by SEQN
aggregated_data <- merged_data %>%
  group_by(SEQN) %>%
  summarise(
    AGE = first(RIDAGEYR),  
    SEX = first(RIAGENDR),
    RACE_ETHNICITY = first(RIDRETH1),
    BMI = first(BMXBMI),
    FAMILY_INCOME = first(INDFMPIR),
    TESTOSTERONE = first(LBXTST),
    ESTROGEN = first(LBXEST),
    SHBG = first(LBXSHBG),
    TOTAL_CALORIES = first(DR1IKCAL),
    SUGAR = first(DR1ISUGR),
    FAT = first(DR1ITFAT),
    PROTEIN = first(DR1IPROT),
    CARBS = first(DR1ICARB)
  ) %>%
  ungroup() 

# Check the result
head(aggregated_data)

```
```{r}

```

```{r}
# Data cleaning and filtering

# Step 1: Age Restriction (10–19 years)
age_filter <- aggregated_data %>%
    filter(AGE >= 10 & AGE <= 19)
ex1 <- nrow(aggregated_data)- nrow(age_filter)
ex1
# Step 2: Exclude Missing Dietary Data (Ultra-Processed Foods)
UPF_filter <- age_filter %>%
    filter(!is.na(TOTAL_CALORIES) & !is.na(SUGAR) & 
           !is.na(FAT) & !is.na(PROTEIN) & 
           !is.na(CARBS))
ex2 <- nrow(age_filter) - nrow(UPF_filter)
ex2
# Step 3: Exclude Missing Hormone Data (Testosterone, Estrogen, SHBG)
hormone_filter <- UPF_filter %>%
    filter(!is.na(ESTROGEN) & !is.na(TESTOSTERONE) & !is.na(SHBG))
ex3 <- nrow(UPF_filter) - nrow(hormone_filter) 
ex3
# Step 4: Exclude Missing Covariate Data
covariate_filter <- hormone_filter %>%
    filter(!is.na(SEX) & !is.na(RACE_ETHNICITY) & !is.na(BMI) &
            !is.na(FAMILY_INCOME))
ex4 <- nrow(hormone_filter) - nrow(covariate_filter) 
ex4
```
```{r}
pacman::p_load(tidyverse, readr, table1, DiagrammeR, rsvg)
# tidyverse is for data management
# readr is for reading in the csv file
# DiagrammeR is for generating Figure 1
# table1 is for generating table 1
# rsvg for exporting figure 1
```
```
```{r}
# Figure 1 to show how the final dataset was arrived at
grViz(diagram = "digraph flowchart{ # gives beginning of flowchart

      # node definitions with substituted label text
      
      node [fontname = Helvetica, shape = rectangle, fontsize=10] 
      
      # Define the nodes: nodes indicate how many boxes you will have in your diagram. Since I have three sets of exclusions, I will have four nodes.
      
      node1 [label = '@@1'] # starting number
      node2 [label = '@@2'] # number after exclusion 1
      node3 [label = '@@3'] # number after exclusion 2
      node4 [label = '@@4'] # number after exclusion 3
      node5 [label = '@@5'] # number after exclusion 4
      # edge definitions with the node IDs are used to indicate how the rectangle boxes flow from each other. 
      
      node1 -> node2 -> node3 -> node4 -> node5
}
      # This set of code provides the text in each rectangle box.
      [1]: 'Aggregated NHANES data 2015-2016  n = 9971'
      [2]: 'Excluding 8230 individuals who are not within 10-19 years  n = 1741'
      [3]: 'Excluding 210 individuals with missing data on Ultra-processed foods  n = 1531'
      [4]: 'Excluding 254 individuals with missing data on TT,E2 and SHBG levels  n = 1277'
      [5]: 'Excluding 117 individuals with missing data on sex,race/ethnicity,BMI and family income  n= 1160'
      ")



```


```{r}
#creating categorical variable 
# step 1 : Creating age categories
# Recoding Age into categories
covariate_filter <- covariate_filter %>%
  mutate(
    AGE_GROUP = case_when(
      AGE >= 10 & AGE <= 12 ~ 1,  # 10-12 years
      AGE >= 13 & AGE <= 16 ~ 2,  # 13-16 years
      AGE >= 17 & AGE <= 19 ~ 3,  # 17-19 years
      TRUE ~ NA_real_
    )) %>%
    
    mutate( SEX = case_when(
      SEX == 1 ~ 1,  # Male
      SEX == 2 ~ 2,  # Female
      TRUE ~ NA_real_
    ),
    
    # Recoding Race/Ethnicity (RIDRETH1)
    RACE_ETHNICITY = case_when(
      RACE_ETHNICITY == 1 ~ 1,  # Non-Hispanic White
      RACE_ETHNICITY == 2 ~ 2,  # Non-Hispanic Black
      RACE_ETHNICITY == 3 ~ 3,  # Mexican American
      RACE_ETHNICITY == 4 ~ 4,  # Other Hispanic
      RACE_ETHNICITY == 5 ~ 5,  # Non-Hispanic Asian
      RACE_ETHNICITY == 6 ~ 6,  # Other/Multiracial
      TRUE ~ NA_real_
    ),
    
    # Recoding BMI (BMXBMI)
    BMI_CATEGORY = case_when(
      BMI < 18.5 ~ 1,  # Underweight
      BMI >= 18.5 & BMI < 25 ~ 2,  # Normal weight
      BMI >= 25 & BMI < 30 ~ 3,  # Overweight
      BMI >= 30 & BMI < 35 ~ 4,  # Obese I
      BMI >= 35 ~ 5,  # Obese II & III
      TRUE ~ NA_real_
    ),
    
    # Recoding Family Income (INDFMPIR) into new categories
    FAMILY_INCOME_CATEGORY = case_when(
      FAMILY_INCOME < 0.50 ~ 1,  # <0.50
      FAMILY_INCOME >= 0.50 & FAMILY_INCOME < 1.00 ~ 2,  # 0.50 - 0.99
      FAMILY_INCOME >= 1.00 & FAMILY_INCOME < 1.24 ~ 3,  # 1.00 - 1.24
      FAMILY_INCOME >= 1.25 & FAMILY_INCOME < 1.49 ~ 4,  # 1.25 - 1.49
      FAMILY_INCOME >= 1.50 & FAMILY_INCOME < 1.84 ~ 5,  # 1.50 - 1.84
      FAMILY_INCOME >= 1.85 & FAMILY_INCOME < 1.99 ~ 6,  # 1.85 - 1.99
      FAMILY_INCOME >= 2.00 ~ 7,  # >2.00
      TRUE ~ NA_real_
    )
  )

# View the updated dataset with all categories
head(covariate_filter)

# Check the distribution of the categories
table(covariate_filter$AGE_GROUP)
table(covariate_filter$SEX)
table(covariate_filter$RACE_ETHNICITY)
table(covariate_filter$BMI_CATEGORY)
table(covariate_filter$FAMILY_INCOME_CATEGORY)


```

```{r}
# Creating table one to show the demographics of the participants
table1(~ as.factor(AGE) + as.factor(SEX) + as.factor(RACE_ETHNICITY) + as.factor(BMI)  + as.factor(FAMILY_INCOME), covariate_filter) 
```


```{r}
#Creating table to show the demographics of the participants stratified by gender
table1(~ as.factor(AGE) + as.factor(RACE_ETHNICITY) + as.factor(BMI)  + as.factor(FAMILY_INCOME)|as.factor(SEX), covariate_filter)


```


```{r}
#Creating table to show the demographics of the participants stratified by BMI
table1(~ as.factor(AGE_GROUP) + as.factor(SEX) + as.factor(RACE_ETHNICITY)  + as.factor(FAMILY_INCOME_CATEGORY)|as.factor(BMI_CATEGORY), covariate_filter)


```


```{r}
install.packages("dplyr")
library(dplyr)

covariate_filter <- covariate_filter %>%
  mutate(FAMILY_INCOME = ifelse(is.na(FAMILY_INCOME), "Unknown", FAMILY_INCOME))

# Ensure it's a factor
covariate_filter$FAMILY_INCOME_CATEGORY <- as.factor(covariate_filter$FAMILY_INCOME)

# Creating table to show the demographics of the participants stratified by Family income to poverty ratio
table1(~ as.factor(AGE) + as.factor(SEX) + as.factor(RACE_ETHNICITY) + as.factor(BMI) | FAMILY_INCOME, covariate_filter)

#i imputed the missing data in family_income.
```

```{r}
# Create a function to classify food based on quantities of macronutrients
categorize_food_by_quantity <- function(CARBS, PROTEIN, FAT, SUGAR, TOTAL_CALORIES) {
  # Handle missing values
  if (is.na(CARBS) || is.na(PROTEIN) || is.na(FAT) || is.na(SUGAR) || is.na(TOTAL_CALORIES)) {
    return("Unknown")
  }
  
  # Ultra-Processed Food (UPF)
  if (SUGAR > 15 || FAT > 20 || TOTAL_CALORIES > 400 || CARBS > 40 || PROTEIN < 5) {
    return("Ultra-Processed Food")
  }
  
  # Processed Food
  else if ((SUGAR >= 5 && SUGAR <= 15) || (FAT >= 5 && FAT <= 20) || (TOTAL_CALORIES >= 150 && TOTAL_CALORIES <= 400) ||
           (CARBS >= 20 && CARBS <= 40) || (PROTEIN >= 5 && PROTEIN <= 10)) {
    return("Processed Food")
  }
  
  # Minimally Processed Food
  else if (SUGAR < 5 && FAT < 5 && TOTAL_CALORIES < 150 && CARBS <= 20 && PROTEIN >= 10) {
    return("Minimally Processed Food")
  }
  
  # If no category matches, label as Unknown
  return("Unknown")
}

# Apply the function to classify each food item
covariate_filter$NOVA_CATEGORY <- mapply(categorize_food_by_quantity, 
                                          covariate_filter$CARBS, 
                                          covariate_filter$PROTEIN, 
                                          covariate_filter$FAT, 
                                          covariate_filter$SUGAR, 
                                          covariate_filter$TOTAL_CALORIES)

# Check the new classification
head(covariate_filter$NOVA_CATEGORY)

summary(covariate_filter)
table(covariate_filter$NOVA_CATEGORY)
head(covariate_filter)


```


```{r}
install.packages("ggplot2")
library(ggplot2)
library(tidyr)
#Creating scatter plots for the hormones and the NOVA food categories
# Reshape data into a long format
long_data <- covariate_filter %>%
  pivot_longer(cols = c(TESTOSTERONE, ESTROGEN, SHBG), 
               names_to = "Hormone", 
               values_to = "Value")

# Scatterplot with facets
ggplot(long_data, aes(x = NOVA_CATEGORY, y = Value, color = NOVA_CATEGORY)) +
  geom_jitter(width = 0.2, height = 0, alpha = 0.6) +
  geom_boxplot(alpha = 0.2, outlier.shape = NA) +
  labs(title = "Hormone Levels by Food Category",
       x = "Food Category",
       y = "Hormone Levels") +
  facet_wrap(~ Hormone, scales = "free_y") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_color_brewer(palette = "Set2")

```

```{r}
library(ggplot2)
#Creating density plots for the hormones and the NOVA food categories
# Density plot for Testosterone
ggplot(covariate_filter, aes(x = TESTOSTERONE, fill = NOVA_CATEGORY)) +
  geom_density(alpha = 0.6) +
  labs(title = "Density of Testosterone Levels by Food Category",
       x = "Testosterone Levels",
       y = "Density") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set2")

# Density plot for Estrogen
ggplot(covariate_filter, aes(x = ESTROGEN, fill = NOVA_CATEGORY)) +
  geom_density(alpha = 0.6) +
  labs(title = "Density of Estrogen Levels by Food Category",
       x = "Estrogen Levels",
       y = "Density") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set3")

# Density plot for SHBG
ggplot(covariate_filter, aes(x = SHBG, fill = NOVA_CATEGORY)) +
  geom_density(alpha = 0.6) +
  labs(title = "Density of SHBG Levels by Food Category",
       x = "SHBG Levels",
       y = "Density") +
  theme_minimal() +
  scale_fill_brewer(palette = "Pastel1")

```


```{r}
# Fit the linear regression model for TESTOSTERONE
model_testosterone <- lm(TESTOSTERONE ~ NOVA_CATEGORY + BMI + AGE + SEX + FAMILY_INCOME, 
                          data = covariate_filter)

# Summary of the model
summary(model_testosterone)

# Confidence intervals for the coefficients
confint(model_testosterone)

# Fit for ESTROGEN
model_estrogen <- lm(log(ESTROGEN) ~ NOVA_CATEGORY + BMI + AGE + SEX + FAMILY_INCOME, 
                     data = covariate_filter)
                     
# Summary of the model
summary(model_estrogen)

# Confidence intervals for the coefficients
confint(model_estrogen)
                     

# Fit for SHBG
model_shbg <- lm(SHBG ~ NOVA_CATEGORY + BMI + AGE + SEX + FAMILY_INCOME, 
                 data = covariate_filter)

```

```{r}
# Install and load the broom package

install.packages("broom")
library(broom)
# Tidy results for TESTOSTERONE
testosterone_results <- tidy(model_testosterone)

# Tidy results for ESTROGEN and SHBG
estrogen_results <- tidy(model_estrogen)
shbg_results <- tidy(model_shbg)

# Combine results into a summary table
results_table <- bind_rows(
  testosterone_results %>% mutate(Outcome = "Testosterone"),
  estrogen_results %>% mutate(Outcome = "Estrogen"),
  shbg_results %>% mutate(Outcome = "SHBG")
)

# Filter for NOVA_CATEGORY and clean up
results_table <- results_table %>%
  filter(term %in% c("NOVA_CATEGORYProcessed Food", "NOVA_CATEGORYUltra-Processed Food")) %>%
  select(Outcome, term, estimate, conf.low, conf.high, p.value) %>%
  rename(
    Exposure = term,
    Coefficient = estimate,
    `Lower 95% CI` = conf.low,
    `Upper 95% CI` = conf.high,
    `P-value` = p.value
  )

# View the final table
print(results_table)

```

```{r}
str(testosterone_results)
str(estrogen_results)
str(shbg_results)

# Fit regression models
testosterone_model <- lm(TESTOSTERONE ~ NOVA_CATEGORY + AGE + SEX + BMI + FAMILY_INCOME, data = covariate_filter)
estrogen_model <- lm(ESTROGEN ~ NOVA_CATEGORY + AGE + SEX + BMI + FAMILY_INCOME, data = covariate_filter)
shbg_model <- lm(SHBG ~ NOVA_CATEGORY + AGE + SEX + BMI + FAMILY_INCOME, data = covariate_filter)

print(testosterone_model)
print(estrogen_model)
print(shbg_model)



```

```{r}
# Add confidence intervals to the regression results
testosterone_results <- broom::tidy(testosterone_model) %>%
  bind_cols(as.data.frame(confint(testosterone_model)) %>% rename(conf.low = `2.5 %`, conf.high = `97.5 %`))

estrogen_results <- broom::tidy(estrogen_model) %>%
  bind_cols(as.data.frame(confint(estrogen_model)) %>% rename(conf.low = `2.5 %`, conf.high = `97.5 %`))

shbg_results <- broom::tidy(shbg_model) %>%
  bind_cols(as.data.frame(confint(shbg_model)) %>% rename(conf.low = `2.5 %`, conf.high = `97.5 %`))

# Combine results into a single table
results_table <- bind_rows(
  testosterone_results %>% mutate(Outcome = "Testosterone"),
  estrogen_results %>% mutate(Outcome = "Estrogen"),
  shbg_results %>% mutate(Outcome = "SHBG")
)

# Filter for the variables of interest
results_table <- results_table %>%
  filter(term %in% c("NOVA_CATEGORYProcessed Food", "NOVA_CATEGORYUltra-Processed Food")) %>%
  select(Outcome, term, estimate, conf.low, conf.high, p.value) %>%
  rename(
    Exposure = term,
    Coefficient = estimate,
    `Lower 95% CI` = conf.low,
    `Upper 95% CI` = conf.high,
    `P-value` = p.value
  )

print(results_table)

```



```{r}
# Linear regression for testosterone with confounder adjustment
testosterone_model <- lm(TESTOSTERONE ~ NOVA_CATEGORY + AGE + SEX + BMI + FAMILY_INCOME, data = covariate_filter)
summary(testosterone_model)

# Confidence intervals for testosterone model
confint(testosterone_model)

# Linear regression for estrogen with confounder adjustment
estrogen_model <- lm(ESTROGEN ~ NOVA_CATEGORY + AGE + SEX + BMI + FAMILY_INCOME, data = covariate_filter)
summary(estrogen_model)

# Confidence intervals for estrogen model
confint(estrogen_model)

# Linear regression for SHBG with confounder adjustment
shbg_model <- lm(SHBG ~ NOVA_CATEGORY + AGE + SEX + BMI + FAMILY_INCOME, data = covariate_filter)
summary(shbg_model)

# Confidence intervals for SHBG model
confint(shbg_model)

```


```{r}
# Model with interaction term
testosterone_interaction <- lm(TESTOSTERONE ~ NOVA_CATEGORY * SEX + AGE + BMI + FAMILY_INCOME_CATEGORY, data = covariate_filter)
summary(testosterone_interaction)

testosterone_bmi_interaction <- lm(TESTOSTERONE ~ NOVA_CATEGORY * BMI_CATEGORY + AGE + SEX + FAMILY_INCOME_CATEGORY, data = covariate_filter)
summary(testosterone_bmi_interaction)

```

```{r}
# Model with interaction between UPF and BMI
testosterone_bmi_interaction <- lm(TESTOSTERONE ~ NOVA_CATEGORY * BMI_CATEGORY + AGE + SEX + FAMILY_INCOME_CATEGORY, data = covariate_filter)

# View summary
summary(testosterone_bmi_interaction)

```

```{r}
library(ggplot2)

# Histogram of the original testosterone distribution
ggplot(covariate_filter, aes(x = TESTOSTERONE)) +
  geom_histogram(bins = 30, fill = "blue", alpha = 0.7) +
  labs(title = "Original Testosterone Distribution", x = "Testosterone", y = "Frequency")

# Repeat for estrogen and SHBG
ggplot(covariate_filter, aes(x = ESTROGEN)) +
  geom_histogram(bins = 30, fill = "green", alpha = 0.7) +
  labs(title = "Original Estrogen Distribution", x = "Estrogen", y = "Frequency")

ggplot(covariate_filter, aes(x = SHBG)) +
  geom_histogram(bins = 30, fill = "purple", alpha = 0.7) +
  labs(title = "Original SHBG Distribution", x = "SHBG", y = "Frequency")




library(ggplot2)
library(tidyr)

# Select the hormone variables and transform to long format
hormone_data <- covariate_filter %>%
  select(TESTOSTERONE, ESTROGEN, SHBG) %>%
  pivot_longer(cols = everything(), names_to = "Hormone", values_to = "Value")

# Plot all distributions together
ggplot(hormone_data, aes(x = Value, fill = Hormone)) +
  geom_histogram(bins = 30, alpha = 0.6, position = "identity") +
  facet_wrap(~ Hormone, scales = "free") +
  scale_fill_manual(values = c("blue", "green", "purple")) +
  labs(
    title = "Distributions of Testosterone, Estrogen, and SHBG",
    x = "Hormone Levels",
    y = "Frequency",
    fill = "Hormone"
  ) +
  theme_minimal()

```

```{r}
# Apply log transformation to the hormones
covariate_filter$LOG_TESTOSTERONE <- log(covariate_filter$TESTOSTERONE)
covariate_filter$LOG_ESTROGEN <- log(covariate_filter$ESTROGEN)
covariate_filter$LOG_SHBG <- log(covariate_filter$SHBG)


library(ggplot2)
library(tidyr)
library(dplyr)

# Create log-transformed variables
covariate_filter <- covariate_filter %>%
  mutate(
    LOG_TESTOSTERONE = log(TESTOSTERONE),
    LOG_ESTROGEN = log(ESTROGEN),
    LOG_SHBG = log(SHBG)
  )

# Combine untransformed and log-transformed variables into long format
hormone_data_combined <- covariate_filter %>%
  select(
    TESTOSTERONE, ESTROGEN, SHBG,
    LOG_TESTOSTERONE, LOG_ESTROGEN, LOG_SHBG
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "Variable",
    values_to = "Value"
  ) %>%
  mutate(
    Transformation = ifelse(substr(Variable, 1, 4) == "LOG_", "Log-Transformed", "Untransformed"),
    Hormone = sub("LOG_", "", Variable)  # Remove "LOG_" prefix
  )

# Plot distributions for both transformations
ggplot(hormone_data_combined, aes(x = Value, fill = Hormone)) +
  geom_histogram(bins = 30, alpha = 0.6, position = "identity") +
  facet_wrap(~ Transformation + Hormone, scales = "free", ncol = 3) +
  scale_fill_manual(values = c("blue", "green", "purple")) +
  labs(
    title = "Distributions of Hormones (Untransformed and Log-Transformed)",
    x = "Hormone Levels",
    y = "Frequency",
    fill = "Hormone"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  )


# Regression models with log-transformed hormones
log_testosterone_model <- lm(LOG_TESTOSTERONE ~ NOVA_CATEGORY + AGE + SEX + BMI + FAMILY_INCOME, data = covariate_filter)
log_estrogen_model <- lm(LOG_ESTROGEN ~ NOVA_CATEGORY + AGE + SEX + BMI + FAMILY_INCOME, data = covariate_filter)
log_shbg_model <- lm(LOG_SHBG ~ NOVA_CATEGORY + AGE + SEX + BMI + FAMILY_INCOME, data = covariate_filter)

# Summarize the models
summary(log_testosterone_model)
summary(log_estrogen_model)
summary(log_shbg_model)


```

```{r}
# Install car package if not already installed
install.packages("car")
library(car)

# Check VIF for the testosterone model
vif(model_testosterone)

# Repeat for other models
vif(model_estrogen)
vif(model_shbg)

# Q-Q plot for residuals of testosterone model
qqnorm(residuals(model_testosterone))
qqline(residuals(model_testosterone), col = "red")

# Shapiro-Wilk test for normality
shapiro.test(residuals(model_testosterone))

# Repeat for other models
qqnorm(residuals(model_estrogen))
qqline(residuals(model_estrogen), col = "red")
shapiro.test(residuals(model_estrogen))

qqnorm(residuals(model_shbg))
qqline(residuals(model_shbg), col = "red")
shapiro.test(residuals(model_shbg))

# Residual plot for testosterone model
plot(fitted(model_testosterone), residuals(model_testosterone),
     main = "Residuals vs Fitted for Testosterone",
     xlab = "Fitted values", ylab = "Residuals")
abline(h = 0, col = "red")

# Repeat for other models
plot(fitted(model_estrogen), residuals(model_estrogen),
     main = "Residuals vs Fitted for Estrogen",
     xlab = "Fitted values", ylab = "Residuals")
abline(h = 0, col = "red")

plot(fitted(model_shbg), residuals(model_shbg),
     main = "Residuals vs Fitted for SHBG",
     xlab = "Fitted values", ylab = "Residuals")
abline(h = 0, col = "red")

# Scale-Location plot for testosterone model
plot(fitted(model_testosterone), sqrt(abs(residuals(model_testosterone))),
     main = "Scale-Location Plot for Testosterone",
     xlab = "Fitted values", ylab = "√|Residuals|")
abline(h = 0, col = "red")

# Breusch-Pagan test for heteroscedasticity
install.packages("lmtest")
library(lmtest)

bptest(model_testosterone)

# Repeat for other models
bptest(model_estrogen)
bptest(model_shbg)



```

```{r}
# Full model for Testosterone
full_model_testosterone <- lm(TESTOSTERONE ~ NOVA_CATEGORY + AGE + SEX + BMI + FAMILY_INCOME, data = covariate_filter)

# Reduced model for Testosterone (excluding FAMILY_INCOME)
reduced_model_testosterone <- lm(TESTOSTERONE ~ NOVA_CATEGORY + AGE + SEX + BMI, data = covariate_filter)

# Likelihood Ratio Test
anova(reduced_model_testosterone, full_model_testosterone)

# Full model for Estrogen
full_model_estrogen <- lm(ESTROGEN ~ NOVA_CATEGORY + AGE + SEX + BMI + FAMILY_INCOME, data = covariate_filter)

# Reduced model for Estrogen (excluding FAMILY_INCOME)
reduced_model_estrogen <- lm(ESTROGEN ~ NOVA_CATEGORY + AGE + SEX + BMI, data = covariate_filter)

# Likelihood Ratio Test
anova(reduced_model_estrogen, full_model_estrogen)

# Full model for SHBG
full_model_shbg <- lm(SHBG ~ NOVA_CATEGORY + AGE + SEX + BMI + FAMILY_INCOME, data = covariate_filter)

# Reduced model for SHBG (excluding FAMILY_INCOME)
reduced_model_shbg <- lm(SHBG ~ NOVA_CATEGORY + AGE + SEX + BMI, data = covariate_filter)

# Likelihood Ratio Test
anova(reduced_model_shbg, full_model_shbg)


```


